---
title: "ISONE DA LMP Futures Settlement Model"
author: Nate Garozzo
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: false
    code-tools: false
    theme: flatly
    embed-resources: true
    fig-width: 8
    fig-height: 5
    fig-cap-location: bottom
    tbl-cap-location: top
    highlight-style: github
    mainfont: "Georgia"
    monofont: "Fira Code"
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
source("~/BestBaseWorstModel/R/run_pipeline.R")
```

```{css}
p code, li code {
  color: #333333 !important;
  background-color: #f5f5f5;
  padding: 2px 4px;
  border-radius: 3px;
}
```

## Introduction

Consumers use ISONE DA-LMP (ISO-New England Day-Ahead locational marginal pricing) futures to hedge forward electricity expenses. However, the futures contract price on any given day provides no extra information about the distribution of expected outcomes for the settled price of the contract. Trained on historical futures and settlement data from 2022-2026, this model produces a predictive distribution for a base case (50th percentile), best case (10th percentile) and worst case (95th percentile) outcome for consumers, given a particular day's futures contract price, settlement month, and time to settlement. Linear regression is used to estimate the conditional mean forecast error and a Gamma GLM to estimate the conditional variance. Finally, a Monte Carlo simulation is layered in to produce the distribution of expected outcomes. The goal for this model is to provide energy consumers in the ISONE territory with a better understanding of their forecasted expenses and whether or not they should be hedging their energy consumption.

## Data

There are three primary data sets used in this analysis. The first is `isone_settled_avg_da_lmp_2013_2026.csv` , which contains the average monthly settled DA-LMP in ISONE for every month from 2013-2026. After data wrangling and tidying, the cleaned data resembles the following sample:

```{r}
monthly_lmps |>
  filter((month == "Jan" & year == "2018") |
           (month == "Jul" & year == "2025")) |>
  knitr::kable(
    col.names = c("Month", "Year", "Settled Avg DA LMP"),
    digits = 2,
  )
```

<br> The next two are `off_peak_isone_historical_futures_2022_2026.csv` and `peak_isone_historical_futures_2022_2026.csv` , which contain the off-peak and peak daily settled futures contract prices (as far as 85 months ahead) from 11/18/2022 to 6/4/2026. For the purposes of training the model, and condensing the data, the off-peak and peak daily futures prices were averaged into a single futures contract price point. Below is another sample:

```{r}
futures |>
  slice(1, 500) |>
  knitr::kable(
    col.names = c("Current Month", "Current Day", "Current Year", 
                  "Delivery Month", "Delivery Year", "Futures Price"),
    digits = 2,
  )
```

<br> Merging these data frames, we can calculate the error ($\text{realized price - futures price}$) for every daily futures price prediction based on its settled monthly average DA-LMP:

```{r}
errors |>
  slice(1, 505) |>
  select(current_month, current_day, current_year, delivery_month, 
         delivery_year, futures_price, error, months_out) |>
  knitr::kable(
    col.names = c("Cur. Month", "Cur. Day", "Cur. Year", "Del. Month",
                  "Del. Year", "Futures Price", "Error", "Months Out"),
    digits = 2,
  )
```

<br> Since the original data sets did not overlap from 2013-2021, there were several observations with no error value. These were subsequently filtered out, as well as any contracts that were either expired or purchased during the delivery month ($\text{months out < 1}$).

## Methodology

Forecasting price ranges poses several challenges. First, observations taken on consecutive days are highly autocorrelated. If the error is large and positive on May 15, the error is also likely to be large and positive on May 16. Rather than modeling the time series of the errors directly, this model takes a different approach, treating each observation as a draw from a distribution defined by its delivery month and time to expiration. It tries to minimize the autocorrelation problem by asking "given a contract with these characteristics, what is the distribution of errors that we have historically observed?"

A second challenge is seasonal variation in price levels and forecast uncertainty. Winter months display much higher volatility than summer months, which is explained largely by the sensitivity of New England electricity prices to natural gas markets. A model which assumes constant volatility for each month of the year would severely misrepresent the risk in several delivery months.

The model has three primary components:

1.  A linear regression estimates the conditional mean forecast error as a function of delivery month and months to expiration

$$\hat{\mu}_{m,H} = \alpha_m + \beta_ms(h)$$

where:

- $\hat{\mu}_{m,h}$ is the predicted mean forecast error for delivery month \$m\$ at horizon \$h\$

- $\alpha_m$ is a month-specific intercept

- $\beta_m$ is a month-specific slope on log time to expiration

- $h$ is months to expiration

- $s(h)$ is a natural cubic spline evaluated at horizon h with 6 degrees of freedom

Taking $s(h)$ as opposed to using $h$ directly is driven by a nonlinear relationship between forecast error and time to maturity. For example, a January contract 1 month out is being priced in December, when weather forecasts are volatile and a cold snap can drive up natural gas demand rapidly. However, a January contract 12 months out has a much more "general winter" expectation and there is plenty of time for the market to adjust to conditions. After fitting various functions to the scatter plot of the relationship between horizon and error, the natural cubic spline with $df = 6$ was selected. Unlike linear, log, and square root transformations, which all assume a monotonically decreasing relationship, the spline flexibility captures the wave-like pattern in the data.

```{r}
errors |>
  group_by(months_out) |>
  summarise(sd_error = sd(error), .groups = "drop") |>
  filter(months_out <= 24) |>
  ggplot(aes(x = months_out, y = sd_error)) +
  geom_point(color = "#2c3e50", size = 2) +
  geom_smooth(method = "lm", se = FALSE,
              aes(color = "Linear"), linetype = "dashed", linewidth = 0.7) +
  geom_smooth(method = "lm", formula = y ~ log(x), se = FALSE,
              aes(color = "Log"), linewidth = 0.7) +
  geom_smooth(method = "lm", formula = y ~ sqrt(x), se = FALSE,
              aes(color = "Square Root"), linewidth = 0.7) +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, df = 6), se = FALSE,
              aes(color = "Spline (df=6)"), linewidth = 0.7) +
  scale_color_manual(values = c(
    "Linear"        = "#e74c3c",
    "Log"           = "#27ae60",
    "Square Root"   = "#e67e22",
    "Spline (df=6)" = "#8e44ad"
  )) +
  labs(
    title    = "Forecast Error Volatility by Time to Maturity",
    x        = "Months to Expiry",
    y        = "Std. Dev. of Error ($/MWh)",
    color    = NULL
  ) +
  theme_minimal(base_family = "Georgia") +
  theme(
    plot.title       = element_text(size = 13, face = "bold"),
    plot.subtitle    = element_text(size = 10, color = "gray40"),
    axis.title       = element_text(size = 10),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )
```

<br> 2. A Gamma Generalized linear model (GLM) is used to estimate the conditional variance of forecast errors as a function of delivery month, months to expiration, **and futures price**

$$\log(\hat{\sigma}^2_{m,h,f}) = \gamma_m + \delta_m s(h) + \phi f$$ where:

- $\hat{\sigma}^2_{m,h,f}$ is the predicted error variance for delivery month $m$, horizon $h$, and futures price $f$

- $\gamma_m$ is a month-specific intercept

- $\delta_m$ is a month-specific slope on log time to expiration

- $\phi$ is the coefficient on futures price $f$

- $s(h)$ is a natural cubic spline evaluated at horizon h with 6 degrees of freedom

- The log link $\log(\hat{\sigma}^2_{m,h,f})$ makes sure that the variance is strictly positive

  Looking at a plot of the squared residuals from the error model ($\text{residual}^2 = (e_{m,h} - \mu_{m,h})^2$), we can see that the distribution roughly follows a Gamma distribution, which is strictly positive and right-skewed.

```{r}
model_data |>
  ggplot(aes(x = residual_sq)) +
  geom_histogram(aes(y = after_stat(density)), bins = 200,
                 fill = "#2980b9", alpha = 0.6) +
  geom_density(fill = "#e74c3c", alpha = 0.2, color = "#e74c3c", linewidth = 0.8) +
  coord_cartesian(xlim = c(0, quantile(model_data$residual_sq, 0.95)),
                  ylim = c(0, 0.003)) +
  labs(
    title = "Distribution of Squared Residuals",
    x     = "Squared Residual",
    y     = "Density"
  ) +
  theme_minimal(base_family = "Georgia") +
  theme(
    plot.title       = element_text(size = 13, face = "bold"),
    axis.title       = element_text(size = 10),
    panel.grid.minor = element_blank()
  )
```

<br> Therefore a Gamma GLM is applied to model the conditional variance of the errors rather than a linear model. Unlike the error model, futures price is included as a predictor. This reflects the observation that forecast uncertainty scales with a contract's price level. For instance, a January futures contract priced at \$150 carries far more volatility than a January contract priced at \$50, even if they have the same time to expiration.

3.  A predictive distribution of standardized residuals, based on the conditional mean $\hat{\mu}_{m,h}$ , conditional variance $\hat{\sigma}^2_{m,h,f}$, and each observed residual $e_i$

    $$z_i = \frac{e_i - \hat{\mu}_{m, h}}{\hat{\sigma}^2_{m,h,f}}$$

where:

- $\hat{\mu}_{m,h}$ is the predicted mean forecast error for delivery month $m$ at horizon $h$

- $e_i$ is the observed forecast error

- $\hat{\mu}_{m, h}$ is the fitted mean error from the error model

- $\hat{\sigma}^2_{m,h,f}$ is the fitted standard deviation from the volatility model

  After adjusting for the conditional mean and scaling by the conditional standard deviation, the residuals should be distributed consistently regardless of the specific month or season. If the standardized residuals were centered well above or below 0 for a particular season, the error model would be systematically over or undepredicting errors for that group. If the spread were drastically wider or narrower for a particular season, then the volatility model would not be capturing the difference in uncertainty across months.

```{r}
model_data |>
  mutate(
    month_group = case_when(
      delivery_month %in% c("Jan", "Feb", "Dec") ~ "Winter",
      delivery_month %in% c("Jun", "Jul", "Aug") ~ "Summer",
      TRUE ~ "Shoulder Months"
    )
  ) |>
  ggplot(aes(x = z)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = "#2980b9", alpha = 0.6) +
  geom_density(color = "#e74c3c", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  facet_wrap(~ month_group) +
  labs(
    title = "Distribution of Standardized Residuals by Season",
    x     = "Standardized Residual (z)",
    y     = "Density"
  ) +
  theme_minimal(base_family = "Georgia") +
  theme(
    plot.title       = element_text(size = 13, face = "bold"),
    axis.title       = element_text(size = 10),
    panel.grid.minor = element_blank(),
    strip.text       = element_text(size = 10, face = "bold")
  )
```

<br> The distribution across seasonal groups appears to be similar, but there are some noticeable differences, including the pronounced peak below 0 in the winter months, as well as the heavier left tails in the shoulder and summer months. It seems that the error and volatility models accounted for most of the differences in forecast errors across delivery months and time to expiration, but some seasonal differences in residual shape remain.

To account for this, the empirical distribution of the residuals is approximated separately depending on the expiration season of the contract. This ensures that the shape of the residual distribution reflects the varying characteristics of summer, winter, and shoulder month errors. For each season, kernel density estimation (KDE) is used to fit a smooth, continuous density curve to the observed values of $z$. The CDF is then derived, mapping each value of $z$ to the probability that a standardized residual falls below it. We invert the CDF to produce a function that maps a random draw $u$ from a Uniform distribution $U(0,1)$ to its corresponding value of $z$.

$$z = F^{-1}(u)$$

A Monte Carlo simulation follows: Take 10,000 uniform draws and pass them through the CDF to get 10,000 draws of $z$ that follow the empirical shape of the standardized residual distribution for that contract's expiration season.

Finally, the simulated settlement price is computed as

$$\tilde{S} = f + \hat{\mu}_{m,h} + \hat{\sigma}_{m,h,f} \cdot z$$ where:

- $\tilde{S}$ is a simulated settlement price
- $f$ is the current futures price
- $\hat{\mu}_{m,h}$ is the predicted mean error from the error model
- $\hat{\sigma}_{m,h,f}$ is the predicted standard deviation from the volatility model
- $z$ is a draw from the seasonal empirical inverse CDF

## Diagnostics

**Coverage**

The first metric used to evaluate the model is coverage. Essentially, coverage encapsulates the proportion of model prediction intervals that contained the month's settled average DA LMP, given a specific month and horizon. For instance, take the expiration month March, and a horizon of 7 months. The model will predict a price distribution for every for every historical observation which matches those criteria, using the futures price at the time. If the realized price falls within the distribution, that observation is covered. The coverage for a particular month/horizon combination is $\frac{\text{\# covered}}{\text{\# total observations}}$ . Since the "Best" outcome cutoff is the 5th percentile, and the "Worst" cutoff is the 95th percentile, we would hope that approximately 90% of historical observations are covered by the model. If a month/horizon combination is "too narrow", than means less than 80% of its historical outcomes were covered. If a month/horizon combination is "too wide", then over 95% of its historical outcomes were covered.

```{r}
q05_winter   <- quantile(model_bundle$inv_cdf_winter(runif(100000)),   0.05)
q05_summer   <- quantile(model_bundle$inv_cdf_summer(runif(100000)),   0.05)
q05_shoulder <- quantile(model_bundle$inv_cdf_shoulder(runif(100000)), 0.05)

q95_winter   <- quantile(model_bundle$inv_cdf_winter(runif(100000)),   0.95)
q95_summer   <- quantile(model_bundle$inv_cdf_summer(runif(100000)),   0.95)
q95_shoulder <- quantile(model_bundle$inv_cdf_shoulder(runif(100000)), 0.95)

errors |>
  mutate(
    season = case_when(
      delivery_month %in% c("Dec", "Jan", "Feb") ~ "winter",
      delivery_month %in% c("Jun", "Jul", "Aug") ~ "summer",
      TRUE                                        ~ "shoulder"
    ),
    mu_hat = predict(error_model, newdata = pick(everything())),
    sd_hat = sqrt(predict(vol_model, newdata = pick(everything()),
                          type = "response")),
    p05 = futures_price + mu_hat + sd_hat * case_when(
      season == "winter"   ~ q05_winter,
      season == "summer"   ~ q05_summer,
      season == "shoulder" ~ q05_shoulder
    ),
    p95 = futures_price + mu_hat + sd_hat * case_when(
      season == "winter"   ~ q95_winter,
      season == "summer"   ~ q95_summer,
      season == "shoulder" ~ q95_shoulder
    ),
    covered = settled_avg_da_lmp >= p05 & settled_avg_da_lmp <= p95,
    horizon = case_when(
      months_out <= 6  ~ "Short (1–6 months)",
      months_out <= 12 ~ "Medium (7–12 months)",
      months_out <= 24 ~ "Long (13–24 months)",
      TRUE             ~ "Very Long (25+ months)"
    )
  ) |>
  group_by(delivery_month, months_out, horizon) |>
  summarise(coverage = mean(covered), .groups = "drop") |>
  group_by(horizon) |>
  summarise(
    `Mean Coverage`  = round(mean(coverage), 3),
    `% Narrow Intervals` = scales::percent(mean(coverage < 0.80), accuracy = 0.1),
    `% Wide Intervals`  = scales::percent(mean(coverage > 0.95), accuracy = 0.1),
    .groups          = "drop"
  ) |>
  arrange(factor(horizon, levels = c(
    "Short (1–6 months)",
    "Medium (7–12 months)",
    "Long (13–24 months)",
    "Very Long (25+ months)"
  ))) |>
  knitr::kable(
    col.names = c("Horizon", "Mean Coverage", "Narrow Intervals", "Wide Intervals"),
    caption   = "Model Calibration by Forecast Horizon (Target: 90%)"
  )
```

Based on the table, all horizon groups seem to have roughly 90% coverage, which is encouraging. The only exception is very long horizons with a mean coverage of 0.853, which makes sense given the distance to settlement. The narrow interval percentage for short, medium, and long horizons ranges from 15.3%-20.8%. This figure is slightly higher than desired, but not too concerning. Wide interval percentage for medium, long, and very long horizons range from 66.7%-67.3%. This is a notable amount of over-coverage, but given the goal of this model, it is better to be overly cautious in predictions rather than too confident.. The wide interval percentage for short horizons is much lower at 40.3%, meaning the model operates with a high degree of precision in shorter time frames.

**Bias**

The next diagnostic is bias, which reveals whether the model is systematically over-or under-predicting in any given delivery month. A T-statistic above 2 or below -2 means the error model has a statistically significant directional bias for that month.

```{r}
errors |>
  mutate(
    mu_hat = predict(error_model, newdata = pick(everything())),
    bias   = error - mu_hat
  ) |>
  group_by(delivery_month) |>
  summarise(
    `Mean Bias`    = round(mean(bias), 3),
    `Std. Dev`     = round(sd(bias), 3),
    `T-Statistic`  = round(mean(bias) / (sd(bias) / sqrt(n())), 3),
    .groups        = "drop"
  ) |>
  arrange(factor(delivery_month, levels = month.abb)) |>
  knitr::kable(
    col.names = c("Delivery Month", "Mean Bias", "Std. Dev.", "T-Statistic"),
    caption   = "Bias Diagnostic by Delivery Month (|T-Stat| > 2 indicates significant bias)"
  )
```

<br>None of the months have a T-statistic above 2 or below -2, so the model passes the bias check.

## Results

**Interpreting the Distributions**

```{r}
library(patchwork)

# Sample contracts
contracts <- tibble(
  delivery_month = c("Jan", "Jul", "Jan"),
  months_out     = c(3, 3, 24),
  futures_price  = c(55, 35, 60),
  label          = c("Jan 3mo", "Jul 3mo", "Jan 24mo")
)

# Run simulations
sims <- contracts |>
  rowwise() |>
  mutate(
    sim = list(forecast_prices(
      model_bundle   = model_bundle,
      delivery_month = delivery_month,
      months_out     = months_out,
      futures_price  = futures_price
    ))
  ) |>
  ungroup()

# Extract simulation draws for plotting
sim_draws <- sims |>
  mutate(draws = map(sim, ~ tibble(price = .x$forecast))) |>
  select(label, draws) |>
  unnest(draws)

# Summary table
summary_table <- sims |>
  mutate(summary = map(sim, summarize_forecast)) |>
  select(summary) |>
  unnest(summary) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  select(delivery_month, months_out, futures_price,
         settlement_volatility, p05_settlement,
         p50_settlement, p95_settlement)

# Distribution plot
p_dist <- sim_draws |>
  mutate(label = factor(label, levels = c("Jan 3mo", "Jul 3mo", "Jan 24mo"))) |>
  ggplot(aes(x = price, fill = label)) +
  geom_density(alpha = 0.5, color = NA) +
  scale_fill_manual(values = c(
    "Jan 3mo"  = "#2980b9",
    "Jul 3mo"  = "#27ae60",
    "Jan 24mo" = "#8e44ad"
  )) +
  labs(
    title    = "Simulated Settlement Price Distributions",
    subtitle = "Winter contracts exhibit wider distributions than summer",
    x        = "Simulated Settlement Price ($/MWh)",
    y        = "Density",
    fill     = NULL
  ) +
  theme_minimal(base_family = "Georgia") +
  theme(
    plot.title       = element_text(size = 13, face = "bold"),
    plot.subtitle    = element_text(size = 10, color = "gray40"),
    axis.title       = element_text(size = 10),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

# Display
p_dist

summary_table |>
  knitr::kable(
    col.names = c("Month", "Months Out", "Futures Price", "Volatility ($/MWh)",
                  "Best Case (p05)", "Base Case (p50)", "Worst Case (p95)"),
    caption   = "Sample Forecast Output"
  )
```

<br> Above is a sample output from the model, with a visual based on its predicted distribution for 3 different contracts and a supporting table. There are some noticeable differences between the distributions. For instance, July 3 months out is much more compact than both January distributions (which makes sense given the heightened uncertainty of prices in winter). January 24 months out is slightly narrower than January 3 months out, which has to do with the near-term volatility of short-horizon winter contracts. January 24 months out is also shifted much further right, displaying how horizon can have a significant impact on the expected error for a contract.

**Assumptions, Limitations, and Future Work**

1.  Assumptions

- The relationship between forecast error, delivery month, and time to expiration is stable over time

- Seasonal groupings sufficiently capture the differences in residual shapes across months

- Monthly average DA LMP cannot be negative (a zero floor is applied to all simulations)

- Peak and off-peak prices are equally weighted in computing the average futures price

2.  Limitations

- The model is trained on data that includes the 2021/2022 winter spike and post-COVID price volatility, which may distort estimates

- No external factors, such as storage levels, gas basis, or weather forecasts are included in the model, limiting its ability to condition on market fundamentals

- The model does not forecast where futures prices will move, only where they will land

3.  Future work

- Expanding the training data set to incorporate a broader range of historical data in the model

- Walk-forward validation to properly assess out-of-sample performance

- Handling of historical extreme winter events in price forecasting
